---
title: "Player performance different metrics"
author: "Anup Walvekar"
output: md_document
---
# Better Economy Rate
Economy Rate (ER) is one of the vary important statistic to evaluate a bowler. The definition of ER is the average number of runs given away by the bowler in one over. This [Cricinfo article](http://www.espncricinfo.com/blogs/content/story/1084584.html) argues how this traditional definition of ER is not a good measure of player quality. The gist of the argument is that the traditional definition ignores the context of the game, i.e. the run rate during a match. So instead of calculating ER as the average number of runs per over, we also need to introduce a way to incorporate the context. They introduce the concept of Economy Rate differential (ERx) in the article, which is the average of player economy rate and innings run rate. There are some problems with this definition:

1. This definition doesn't take into account the number of balls the player bowled in a game
2. The value of this statistics are very different than the traditional statistic (ER), so it could be hard to comprehend.
3. ERx compares players of same team well, but it fails to encorporate the absolute value of players abilities. e.g. if an average player is bowling with bad bowlers, his ERx stat is going to be very good, may be even better than best of the bowlers.

### New Economy Rate definitions

To bypass these problems I defined following economy rating definitions.

1. ER1: 6 x runs/balls,
    This is the traditional definition of economy rates that is average number of runs per over.
2. ER2: 6 x sum(runs_i x rr_i)/sum(balls_i x rr_i), where rr_i is run-rate of ith match,
    This definition captures the context of each game that bowler plays. It puts higher weight on the matches with higher run rates.
3. ER3: ER1 + sum(w_i x (rr_i - eco_i)), where eco_i is economy rate of the bowler in ith match.
    This definition has two components the first component is ER1, that is the individual level of player where as the second component tries to capture the context of matches when bowler has bowled. The weights w_i are proportional to the number of balls bowled by the bowler in ith match.
4. R_i is the ranking of players w.r.t. ERi statistics.

I rank all players who have bowled at least 15 overs (60 balls) in a given season.



```{r read_data,echo=FALSE,include=FALSE}
library(data.table)
library(knitr)
events = read.csv('../../CSV/events.csv',sep = ';',header = T)
match_info = read.csv('../../CSV/match_info.csv',sep = ';',header = T)
match_info$season = substr(match_info$start_date,1,4)
events = merge(match_info[,c('match_id','season')]
               ,events
               ,by = 'match_id')
events = as.data.table(events)
# head(events)
innings_summary = events[,list(match_balls = length(ball[!(extra_type %in%
                                                              c('wides'
                                                                ,'noballs'))])
                               ,match_runs = sum(total_runs)
                               ,match_extra_runs = sum(extra_runs)
                               ,match_wickets = length(out_type[out_type != 'none']))
                         ,by="season,match_id,innings,batting_team,bowling_team"]
# kable(head(innings_summary))
bowler_summary = events[,list(runs=sum(total_runs[!(extra_type %in% c("byes","legbyes"))])
                              ,balls=length(extra_type[!(extra_type %in% c('noballs','wides'))])
                              ,wickets=length(out_type[out_type%in%c('bowled'
                                                                     ,'caught'
                                                                     ,'caught and bowled'
                                                                     ,'hit wicket'
                                                                     ,'lbw'
                                                                     ,'stumped')]))
                        ,by="season,match_id,innings,batting_team,bowling_team,bowler"]
bowler_summary = merge(bowler_summary
                       ,innings_summary[,c('season'
                                           ,'match_id'
                                           ,'innings'
                                           ,'match_balls'
                                           ,'match_runs'
                                           ,'match_wickets')]
                       ,by = c('season','match_id','innings'))
bowler_summary$player_eco = 6*bowler_summary$runs/bowler_summary$balls
bowler_summary$inning_rr = 6*bowler_summary$match_runs/bowler_summary$match_balls
bowler_summary$inning_sr = ifelse(bowler_summary$match_wickets > 0
                                  ,bowler_summary$match_balls/ bowler_summary$match_wickets
                                  ,0) # 0 is good for later computation
bowler_eco_sr = bowler_summary[,list(matches=length(unique(match_id))
                                     ,balls = sum(balls)
                                     ,runs = sum(runs)
                                     ,wickets = sum(wickets)
                                     ,ER1=6.0*sum(runs)/sum(balls)
                                      ,ER2=6.0*sum(runs*inning_rr)/ sum(balls*inning_rr)
                                     ,ER3 = 6.0*sum(runs)/sum(balls) +
                                       sum((inning_rr-player_eco)*balls)/sum(balls))
                                ,by="season,bowler"]
bowler_eco_sr = subset(bowler_eco_sr,balls > 60)
bowler_eco_sr[,R_1:=rank(ER1,ties.method = "first")
              ,by = "season"]
bowler_eco_sr[,R_2:=rank(ER2,ties.method = "first")
              ,by = "season"]
bowler_eco_sr[,R_3:=rank(ER3,ties.method = "first")
              ,by = "season"]
bowler_eco_sr = bowler_eco_sr[order(season,R_1,R_2,R_3)]
```

The data contains information about all deliveries for all the `r length(unique(events$match_id))` matches from past `r length(unique(events$season))` IPL seasons.

The resuls are as follows.

### Season 2008

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2008),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2009

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2009),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2010

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2010),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2011

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2011),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2012

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2012),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2013

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2013),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2014

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2014),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2015

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2015),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

### Season 2016

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2016),c("bowler"
                                                         ,"balls"
                                                         ,"runs"
                                                         ,"wickets"
                                                         ,"ER1"
                                                         ,"R_1"
                                                         ,"ER2"
                                                         ,"R_2"
                                                         ,"ER3"
                                                         ,"R_3")],n = 10))
```

