---
title: "Player performance different metrics"
author: "Anup Walvekar"
output:  
  html_document:  
        keep_md: true
---
# Better Economy Rate
Economy Rate (ER) is one of the vary important statistic to evaluate a bowler. The definition of ER is the average number of runs given away by the bowler in one over. This [Cricinfo article](http://www.espncricinfo.com/blogs/content/story/1084584.html) argues how this traditional definition of ER is not a good measure of player quality. The gist of the argument is that the traditional definition ignores the context of the game, i.e. the run rate during a match. So instead of calculating ER as the average number of runs per over, we also need to introduce a way to incorporate the context. They introduce the concept of Economy Rate differential (ERx) in the article, which is the average of player economy rate and innings run rate. There are couple of problems with this definition:

1. This definition doesn't take into account the number of balls the player bowled in a game
2. The value of this statistics are very different than the traditional statistic (ER), so it could be hard to comprehend.

To bypass these problems I defined a Weighted Economy Rate (WER), where weights are based on the number of ball bowled by player and the overall run rate of the inning.

$$ \mbox{WER = 6.0 x sum(ER x balls x RR)/sum(balls x RR)} $$
Lets see if this changes anything.

```{r read_data,echo=FALSE}
library(data.table)
library(knitr)
events = read.csv('../CSV/events.csv',sep = ';',header = T)
match_info = read.csv('../CSV/match_info.csv',sep = ';',header = T)
match_info$season = substr(match_info$start_date,1,4)
events = merge(match_info[,c('match_id','season')]
               ,events
               ,by = 'match_id')
```

The events data table has information about all deliveries for `r length(unique(events$match_id))` matches from past `r length(unique(events$season))` IPLs.

```{r}
events = as.data.table(events)
head(events)
```

We need run rate for all innings in order to get weighted economy rate.
```{r}
innings_summary = events[,list(match_balls = length(ball[!(extra_type %in%
                                                              c('wides'
                                                                ,'noballs'))])
                               ,match_runs = sum(total_runs)
                               ,match_extra_runs = sum(extra_runs)
                               ,match_wickets = length(out_type[out_type != 'none']))
                         ,by="season,match_id,innings,batting_team,bowling_team"]
kable(head(innings_summary))
```




```{r}
bowler_summary = events[,list(runs=sum(total_runs[!(extra_type %in% c("byes","legbyes"))])
                              ,balls=length(extra_type[!(extra_type %in% c('noballs','wides'))])
                              ,wickets=length(out_type[out_type%in%c('bowled'
                                                                     ,'caught'
                                                                     ,'caught and bowled'
                                                                     ,'hit wicket'
                                                                     ,'lbw'
                                                                     ,'stumped')]))
                        ,by="season,match_id,innings,batting_team,bowling_team,bowler"]
bowler_summary = merge(bowler_summary
                       ,innings_summary[,c('season'
                                           ,'match_id'
                                           ,'innings'
                                           ,'match_balls'
                                           ,'match_runs'
                                           ,'match_wickets')]
                       ,by = c('season','match_id','innings'))
bowler_summary$inning_rr = 6*bowler_summary$match_runs/bowler_summary$match_balls
bowler_summary$inning_sr = ifelse(bowler_summary$match_wickets > 0
                                  ,bowler_summary$match_balls/ bowler_summary$match_wickets
                                  ,0) # 0 is good for later computation
bowler_eco_sr = bowler_summary[,list(matches=length(unique(match_id))
                                     ,balls = sum(balls)
                                     ,runs = sum(runs)
                                     ,wickets = sum(wickets)
                                     ,regular_eco=6.0*sum(runs)/sum(balls)
                                      ,weighted_eco=6.0*sum(runs*inning_rr)/ sum(balls*inning_rr)
                                      ,regular_sr=sum(balls)/sum(wickets)
                                      ,weighted_sr=sum(balls*inning_sr)/ sum(wickets*inning_sr))
                                ,by="season,bowler"]
bowler_eco_sr = subset(bowler_eco_sr,balls > 60)
bowler_eco_sr[,reg_rank:=rank(regular_eco,ties.method = "first")
              ,by = "season"]
bowler_eco_sr[,wt_rank:=rank(weighted_eco,ties.method = "first")
              ,by = "season"]
bowler_eco_sr = bowler_eco_sr[order(season,reg_rank,wt_rank)]
```

### Season 2008

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2008),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2009

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2009),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2010

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2010),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2011

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2011),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2012

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2012),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2013

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2013),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2014

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2014),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2015

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2015),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

### Season 2016

```{r,echo=FALSE}
kable(head(bowler_eco_sr[which(bowler_eco_sr$season == 2016),c("bowler"
                                                         ,"balls"
                                                          ,"regular_eco"
                                                          ,"reg_rank"
                                                          ,"weighted_eco"
                                                          ,"wt_rank")],n = 10))
```

